Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-31 doc/thesis/appendix.Rnw
Reorganizing

>>>>

\chapter{Mathematical models, likelihood, and Bayesian inference}
\label{chp:prelim}

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $x$ and outcomes $y$. A
\defn{parametric} model defines a family of possible relationships between data
and outcomes, parameterized by one or more numeric parameters $\theta$. A
\defn{statistical} model describes the relationship between data and outcomes
in terms of probabilities. Statistical models define, either explicitly or
implicitly, the probability of observing $y$ given $x$ and, if the model is
parametric, $\theta$. Note that it is entirely possible to have no data $x$,
only observed outcomes $y$. In this case, a model would describe the process by
which $y$ is generated.

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
are vectors of real numbers. The linear model postulates that the outcomes are
linearly related to the data, modulo some noise introduced by measurement
error, environmental fluctuations, and other external factors. Formally, $y_i =
\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
We can make this model a statistical one by hypothesizing a distribution for
the error terms $\varepsilon_i$; most commonly, it is assumed that they are
normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
can see from this formulation that the model is parametric, with parameters
$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
density $\pi$ of observing outcome $y_i$ given the parameters,
\[
  \pi(y \mid \beta, \sigma) = 
  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
\]
where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
distribution with mean zero and variance $\sigma^2$. Note that we are treating
the $x_i$ as fixed quantities and therefore have not conditioned the
probability density on $x$. Also, we have assumed that all the $y_i$ are
independent.

For a general model, the probability density of $y$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
The higher the value of the likelihood, the more likely the observations $y$
are under the model. Thus, the likelihood provides a natural criterion for
fitting the model parameters: we want to pick $\theta$ such that the
probability density of our observed outcomes $y$ is as high as possible. The
parameters that optimize the likelihood are known as the \textit{\gls{ML}}
estimates, denoted $\hat{\theta}$. That is,
\[
  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
\]
\Gls{ML} estimation is usually performed with numerical optimization. In the
simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
y)$ is calculated for each, and the parameters that produce the highest value
are accepted. Many sophisticated numerical optimization methods exist, although
they may not be guaranteed to find the true \gls{ML} estimates if the
likelihood function is multi-modal.

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or belief about what $\theta$ are
likely to be. For example, in the linear regression case, the instrument used
to measure the outcomes may have a well-known margin of error, or the sign of
the slope may be obvious from previous experiments. The Bayesian approach to
model fitting makes use of this information by codifying the investigator's
beliefs as a \defn{prior distribution} on the parameters, denoted
$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
focuses on the product of the likelihood and the prior, $f(y \mid \theta)
\pi(\theta)$. Bayes' theorem tells us that this product is related to the
\textit{posterior distribution} on $\theta$,
\begin{align}
  f(\theta \mid y) 
    = \frac{f(y \mid \theta) \pi(\theta)}
           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
  \label{eq:bayes}
\end{align}
In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
The resulting optimal parameters are called the \gls{MAP} estimates. However,
from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
but rather a random variable with an associated distribution (the posterior).
Therefore, the \gls{MAP} estimate by itself is of limited value without
associated statistics about the posterior distribution, such as the mean or
credible intervals. Unfortunately, to calculate such statistics, it is
necessary to evaluate the normalizing constant in the denominator of
\cref{eq:bayes}, which is almost always an intractable integral.

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain {\color{red}\sout{whose states are indexed by
possible model parameters. The transition probability from state $\theta_1$ to
state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
model parameters. The transition density from parameters $\theta_1$ to
$\theta_2$ is taken to be}}
\[
  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and can be accomplished via the Metropolis-Hastings
algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).

\begin{algorithm}
  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
  \begin{algorithmic}
    \State Draw $\theta$ according to the prior $\pi(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\min \left( 1, 
       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

### 2016-05-31 doc/thesis/intro.tex
Reorganizing

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
  \label{eq:post}
\end{align}
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will approximate it using a \defn{Monte Carlo} approach.
The fundamental idea behind Monte Carlo methods is succinctly expressed by
\textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions, including as the aforementioned
posterior, that are analytically intractable and from which direct sampling is
not possible } (for a review see \autocite{robert2004monte}).
\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte} \uline{is one such algorithm.}

@@

\glsreset{ABC}

\subsection{Model fitting}
\label{subsec:mfit}

A \defn{mathematical model} is a formal description of a hypothesized
relationship between some observed data, $x$ and outcomes $y$. A
\defn{parametric} model defines a family of possible relationships between data
and outcomes, parameterized by one or more numeric parameters $\theta$. A
\defn{statistical} model describes the relationship between data and outcomes
in terms of probabilities. Statistical models define, either explicitly or
implicitly, the probability of observing $y$ given $x$ and, if the model is
parametric, $\theta$. Note that it is entirely possible to have no data $x$,
only observed outcomes $y$. In this case, a model would describe the process by
which $y$ is generated.

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
are vectors of real numbers. The linear model postulates that the outcomes are
linearly related to the data, modulo some noise introduced by measurement
error, environmental fluctuations, and other external factors. Formally, $y_i =
\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
We can make this model a statistical one by hypothesizing a distribution for
the error terms $\varepsilon_i$; most commonly, it is assumed that they are
normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
can see from this formulation that the model is parametric, with parameters
$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
density $\pi$ of observing outcome $y_i$ given the parameters,
\[
  \pi(y \mid \beta, \sigma) = 
  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
\]
where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
distribution with mean zero and variance $\sigma^2$. Note that we are treating
the $x_i$ as fixed quantities and therefore have not conditioned the
probability density on $x$. Also, we have assumed that all the $y_i$ are
independent.

For a general model, the probability density of $y$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
The higher the value of the likelihood, the more likely the observations $y$
are under the model. Thus, the likelihood provides a natural criterion for
fitting the model parameters: we want to pick $\theta$ such that the
probability density of our observed outcomes $y$ is as high as possible. The
parameters that optimize the likelihood are known as the \textit{\gls{ML}}
estimates, denoted $\hat{\theta}$. That is,
\[
  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
\]
\Gls{ML} estimation is usually performed with numerical optimization. In the
simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
y)$ is calculated for each, and the parameters that produce the highest value
are accepted. Many sophisticated numerical optimization methods exist, although
they may not be guaranteed to find the true \gls{ML} estimates if the
likelihood function is multi-modal.

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or belief about what $\theta$ are
likely to be. For example, in the linear regression case, the instrument used
to measure the outcomes may have a well-known margin of error, or the sign of
the slope may be obvious from previous experiments. The Bayesian approach to
model fitting makes use of this information by codifying the investigator's
beliefs as a \defn{prior distribution} on the parameters, denoted
$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
focuses on the product of the likelihood and the prior, $f(y \mid \theta)
\pi(\theta)$. Bayes' theorem tells us that this product is related to the
\textit{posterior distribution} on $\theta$,
\begin{align}
  f(\theta \mid y) 
    = \frac{f(y \mid \theta) \pi(\theta)}
           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
  \label{eq:bayes}
\end{align}
In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
The resulting optimal parameters are called the \gls{MAP} estimates. However,
from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
but rather a random variable with an associated distribution (the posterior).
Therefore, the \gls{MAP} estimate by itself is of limited value without
associated statistics about the posterior distribution, such as the mean or
credible intervals. Unfortunately, to calculate such statistics, it is
necessary to evaluate the normalizing constant in the denominator of
\cref{eq:bayes}, which is almost always an intractable integral.

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain {\color{red}\sout{whose states are indexed by
possible model parameters. The transition probability from state $\theta_1$ to
state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
model parameters. The transition density from parameters $\theta_1$ to
$\theta_2$ is taken to be}}
\[
  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and can be accomplished via the Metropolis-Hastings
algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).

\begin{algorithm}
  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
  \begin{algorithmic}
    \State Draw $\theta$ according to the prior $\pi(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\min \left( 1, 
       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

@@

%\subsection{Convergence properties}
%
%The convergence of \gls{ABC} to the true posterior distribution of interest
%depends on two components: the convergence of the underlying algorithm to the
%\gls{ABC} target distribution, and the resemblance of this target distribution
%to the true
%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
%address these components one at a time. Rejection \gls{ABC} and
%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
%sample from the target distributions. Both of these methods are
%well-established and their convergence properties have been extensively
%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
%}restrict our attention to the convergence properties of \gls{SMC}.

@@

{\color{blue}\uline{ Our method integrates the four distinct research areas
  just described: phylogenetics, contact networks, \acrlong{SMC}, and
  \acrlong{ABC}. The first two topics together form the problem domain.
  Phylogenetic data is the input to our method, while estimates of the
  parameters of contact network models are the desired output. The latter two
  topics define the algorithm and statistical framework that our inference
method will use. }}

>>>>

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the \emph{posterior distribution}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
  \label{eq:post}
\end{align}
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will approximate it using a \defn{Monte Carlo} approach.
The fundamental idea behind Monte Carlo methods is succinctly expressed by
\textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions, including as the aforementioned
posterior, that are analytically intractable and from which direct sampling is
not possible } (for a review see \autocite{robert2004monte}).
\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte} \uline{is one such algorithm.}

@@

\glsreset{ABC}
{\color{blue}\uline{\acrlong{SMC}, and the \gls{SMC} sampler, were developed
for sampling from distributions which can be evaluated up to a normalizing
constant. TODO }

\uline{This section is organized as follows. First, we formalize the concept of a
mathematical model and discuss what it means to ``fit'' such models to data 
and how this is typically accomplished. Second, we introduce \gls{ABC}, which
was designed for situations like the one we face in this work where the most
commonly used means of fitting models cannot be applied. Finally, we describe
algorithms which can be used to implement \gls{ABC}, including a variant of
\gls{SMC}.}}

@@

@@

{\color{blue}\uline{ Our method integrates the four distinct research areas
  just described: phylogenetics, contact networks, \acrlong{SMC}, and
  \acrlong{ABC}. The first two topics together form the problem domain.
  Phylogenetic data is the input to our method, while estimates of the
  parameters of contact network models are the desired output. The latter two
  topics define the algorithm and statistical framework that our inference
method will use. }

\uline{Having introduced transmission trees and contact network models, we are
now in a position to argue that approximate Bayesian computation is an
appropriate tool to apply towards our main objective: fitting network models to
phylogenetic data. Let $M$ be a contact network model having parameters
$\theta$.

}}

### 2016-05-31 doc/thesis/methods_results.Rnw
Reorganizing

{\color{blue}\uline{
The chapter is organized as follows. One section is devoted to each of our
three research aims: the development of a method for inference of contact
network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
investigation of the \acrlong{BA} network model using simulated data
}(\cref{sec:ba}), \uline{ and the application of the method to real-world
\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
description of the methods used and, in the case of the latter two objectives,
the results of the experiments. The fourth and final section of the chapter
}(\cref{sec:disc}) \uline{contains a unified discussion of all three
objectives. }}

@@

>>>>

{\color{blue}\uline{
This chapter is organized as follows. One section is devoted to each of our
three research aims: the development of a method for inference of contact
network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
investigation of the \acrlong{BA} network model using simulated data
}(\cref{sec:ba}), \uline{ and the application of the method to real-world
\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
description of the methods used and, in the case of the latter two objectives,
the results of the experiments. The fourth and final section of the chapter
}(\cref{sec:disc}) \uline{contains a unified discussion of all three
objectives. }}

@@

\subsection{Justification for approach}

We present here a mathematical justification for the use of \gls{ABC} for the
problem at hand, as opposed to more frequently-used approaches for fitting
mathematical models. Consider a contact network model with parameters $\theta$,
and an estimated transmission tree $T$. TODO

### 2016-05-31 doc/thesis/objective.tex
Reorganizing

@@

{\color{blue}\uline{
In a Bayesian setting, our goal is to estimate the posterior distribution on
model parameters given our data, $\Pr(\theta \mid T)$. Bayes' theorem relates
the posterior distribution to the likelihood,}
\[
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
\]
\uline{For all but the simplest mathematical models, the calculation of the
normalizing constant $\Pr(T)$ is an intractable problem. What we have argued in
the previous paragraph is that calculating $\Pr(T \mid \theta)$ for a single
parameter candidate $\theta$ also seems likely to be computationally hard.
Therefore, the problem of fitting contact network models to phylogenies seems
to be of the \emph{doubly-intractible} type~\autocite{murray2012mcmc}, which
would imply that these models are not amenable to neither \gls{ML} nor Bayesian
inference techniques. Although both methods are able to cope with an
intractible normalizing constant (for example, by local search for \gls{ML} or
Bayesian \gls{MCMC}), neither can avoid the intractable likelihood
calculations. }}

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring,
fu1997estimating, beaumont2002approximate}. All of the ingredients required to
apply \gls{ABC} to this problem are readily available. Simulating networks is
straightforward under a variety of models. Epidemics on those networks, and the
corresponding transmission trees, can also be easily simulated. As mentioned
above, contact networks can profoundly affect transmission tree shape. Those
shapes can be compared using a highly informative similarity measure called the
``tree kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
functions have been demonstrated to work well as distance functions in
\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with \gls{SMC},
which has several advantages over other
algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
even more attractive approach~\autocite{del2012adaptive}. In summary, our
method to infer contact network parameters will combine the following:
stochastic simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on the
framework developed by~\autocite{poon2015phylodynamic}, who combined \gls{ABC}
with the tree kernel to infer parameters of population genetic models from
viral phylogenies using \gls{MCMC}. }}

>>>>

{\color{blue}\uline{In this work, we take a Bayesian approach: our goal is to
estimate the posterior distribution on model parameters given our data,}
\[
    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
\]
\uline{where $\Pr(T \mid \theta)$ is the likelihood, $\Pr(\theta)$ is the prior,
and $\Pr(T)$ is the marginal probability of $T$ which acts as a normalizing
constant on the posterior (see \cref{chp:prelim} \uline{for a review of
mathematical modeling and Bayesian inference, including definitions of these
concepts).
}}

@@

{\color{blue}\uline{
\uline{
For all but the simplest mathematical models, the calculation of the
normalizing constant $\Pr(T)$ is an intractable problem. 
What we have argued in
the previous paragraph is that calculating $\Pr(T \mid \theta)$ for a single
parameter candidate $\theta$ also seems likely to be computationally hard.
Therefore, the problem of fitting contact network models to phylogenies seems
to be of the \emph{doubly-intractible} type~\autocite{murray2012mcmc}, which
would imply that these models are not amenable to neither \gls{ML} nor Bayesian
inference techniques. Although both methods are able to cope with an
intractible normalizing constant (for example, by local search for \gls{ML} or
Bayesian \gls{MCMC}), neither can avoid the intractable likelihood
calculations. }}

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring,
fu1997estimating, beaumont2002approximate}. All of the ingredients required to
apply \gls{ABC} to this problem are readily available. Simulating networks is
straightforward under a variety of models. Epidemics on those networks, and the
corresponding transmission trees, can also be easily simulated. As mentioned
above, contact networks can profoundly affect transmission tree shape. Those
shapes can be compared using a highly informative similarity measure called the
``tree kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
functions have been demonstrated to work well as distance functions in
\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with several
algorithms, but \gls{SMC} has advantages over others,
{\color{blue}\uline{including improved accuracy in low-density regions and
parallelizability}}~\autocite{mckinley2009inference}. A recently-developed
adaptive algorithm requiring minimal tuning on the part of the user makes
\gls{SMC} an even more attractive approach~\autocite{del2012adaptive}. In
summary, our method to infer contact network parameters will combine the
following: stochastic simulation of epidemics on networks, the tree kernel, and
adaptive \gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on
the framework developed by~\autocite{poon2015phylodynamic}, who combined
\gls{ABC} with the tree kernel to infer parameters of population genetic models
from viral phylogenies using \gls{MCMC}. }}
